---
title: "Arcas-HLA Typing Report"
author: "Foo Cheung"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
library(jsonlite)
library(tidyverse)
library(knitr)
library(kableExtra)

input_dir <- "arcasHLA_results"
out_dir <- "arcasHLA_report_outputs"

dir.create(out_dir, showWarnings = FALSE)

files <- list.files(
  input_dir,
  pattern = "genotype.json$",
  recursive = TRUE,
  full.names = TRUE
)

stopifnot(length(files) > 0)
```

# Overview

Arcas-HLA was used to infer HLA genotypes from paired-end bulk RNA-seq FASTQ files. The analysis focused on the following HLA loci:

```{r}
genes_requested <- c("A", "B", "C", "DQA1", "DQB1", "DRB1")
genes_requested
```

These results are intended to support downstream interpretation of SARS-CoV-2-responsive TCRs. Since the published SARS-CoV-2 TCR databases are primarily focused on CD8 T cells, the most important HLA loci for interpretation are HLA-A, HLA-B, and HLA-C.

# Load Arcas-HLA genotype calls

```{r}
hla_df <- map_dfr(files, function(f) {
  sample <- basename(dirname(f))
  dat <- fromJSON(f)

  map_dfr(names(dat), function(gene) {
    alleles <- dat[[gene]]

    tibble(
      Sample = sample,
      Gene = gene,
      Allele1 = ifelse(length(alleles) >= 1, alleles[1], NA),
      Allele2 = ifelse(length(alleles) >= 2, alleles[2], NA),
      Genotype = paste(
        ifelse(length(alleles) >= 1, alleles[1], NA),
        ifelse(length(alleles) >= 2, alleles[2], NA),
        sep = " / "
      )
    )
  })
})

hla_df <- hla_df %>%
  mutate(
    Sample = str_replace(Sample, "_R1_001$", ""),
    Gene = factor(Gene, levels = genes_requested)
  ) %>%
  arrange(Sample, Gene)

write_csv(hla_df, file.path(out_dir, "arcasHLA_genotype_long.csv"))
```

# HLA genotype table

```{r}
hla_wide <- hla_df %>%
  select(Sample, Gene, Genotype) %>%
  pivot_wider(
    names_from = Gene,
    values_from = Genotype,
    names_prefix = "HLA_"
  ) %>%
  arrange(Sample)

write_csv(hla_wide, file.path(out_dir, "arcasHLA_genotype_wide.csv"))

hla_wide %>%
  kable() %>%
  kable_styling(full_width = FALSE)
```

# Class I HLA calls for CD8 TCR interpretation

```{r}
hla_classI <- hla_wide %>%
  select(Sample, HLA_A, HLA_B, HLA_C)

write_csv(hla_classI, file.path(out_dir, "arcasHLA_classI_for_CD8_TCR.csv"))

hla_classI %>%
  kable() %>%
  kable_styling(full_width = FALSE)
```

# Genotype heatmap

```{r}
# Better genotype table figure

hla_plot_table <- hla_df %>%
  mutate(
    Genotype_short = Genotype %>%
      str_replace_all(":01(?=\\s|$| /)", "") %>%
      str_replace_all(":01(?= /)", "")
  ) %>%
  select(Sample, Gene, Genotype_short) %>%
  pivot_wider(
    names_from = Gene,
    values_from = Genotype_short,
    names_prefix = "HLA_"
  ) %>%
  arrange(Sample)

hla_plot_table %>%
  kable(
    caption = "Arcas-HLA genotype calls by sample",
    align = "l"
  ) %>%
  kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "hover", "condensed")
  ) %>%
  scroll_box(width = "100%", height = "500px")

write_csv(
  hla_plot_table,
  file.path(out_dir, "arcasHLA_genotype_table_for_report.csv")
)
```

# Allele frequency summary

```{r}
hla_long <- hla_df %>%
  pivot_longer(
    cols = c(Allele1, Allele2),
    names_to = "Allele_Copy",
    values_to = "Allele"
  ) %>%
  filter(!is.na(Allele), Allele != "")

allele_counts <- hla_long %>%
  count(Gene, Allele, sort = TRUE)

write_csv(allele_counts, file.path(out_dir, "arcasHLA_allele_counts.csv"))

allele_counts %>%
  kable() %>%
  kable_styling(full_width = FALSE)
```

```{r}
p_counts <- ggplot(
  allele_counts,
  aes(x = reorder(Allele, n), y = n)
) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ Gene, scales = "free_y") +
  theme_bw() +
  labs(
    title = "HLA allele counts across samples",
    x = "Allele",
    y = "Count"
  )

p_counts

ggsave(
  file.path(out_dir, "arcasHLA_allele_counts.png"),
  p_counts,
  width = 12,
  height = 8,
  dpi = 300
)

ggsave(
  file.path(out_dir, "arcasHLA_allele_counts.pdf"),
  p_counts,
  width = 12,
  height = 8
)
```

# Two-field simplified HLA calls

For downstream TCR restriction matching, two-field resolution is often easier to use. For example, `A*02:01:01` becomes `A*02:01`.

```{r}
simplify_hla <- function(x) {
  ifelse(
    is.na(x),
    NA,
    str_extract(x, "^[A-Z0-9]+\\*\\d+:\\d+")
  )
}

hla_simple <- hla_df %>%
  mutate(
    Allele1_2field = simplify_hla(Allele1),
    Allele2_2field = simplify_hla(Allele2),
    Genotype_2field = paste(Allele1_2field, Allele2_2field, sep = " / ")
  )

hla_simple_wide <- hla_simple %>%
  select(Sample, Gene, Genotype_2field) %>%
  pivot_wider(
    names_from = Gene,
    values_from = Genotype_2field,
    names_prefix = "HLA_"
  ) %>%
  arrange(Sample)

write_csv(hla_simple_wide, file.path(out_dir, "arcasHLA_genotype_2field_wide.csv"))

hla_simple_wide %>%
  kable() %>%
  kable_styling(full_width = FALSE)
```

# T22 HLA-A call inspection

One sample, `Manthiram_T22_S44`, was called as:

```{r}
hla_wide %>%
  filter(Sample == "Manthiram_T22_S44") %>%
  select(Sample, HLA_A, HLA_B, HLA_C) %>%
  kable() %>%
  kable_styling(full_width = FALSE)
```

The HLA-A call includes `A*02:844`, which is less common than alleles such as `A*02:01`. Review of the Arcas-HLA genotype log showed that the sample had strong HLA-A read support and that multiple closely related A*02 alleles had similar support. The most important biological interpretation for the SARS-CoV-2 TCR analysis is that this sample carries an HLA-A*02 family allele and an HLA-A*31 allele.

# Summary

The Arcas-HLA run successfully generated complete HLA calls for all samples across the requested loci. The Class I calls, especially HLA-A, HLA-B, and HLA-C, are the most important results for interpreting SARS-CoV-2-responsive CD8 TCRs. These HLA calls can be used to evaluate whether published TCR-HLA restrictions are compatible with each subject's inferred HLA genotype.

Key output files generated by this report:

```{r}
list.files(out_dir, full.names = FALSE)
```

